---
title: "cm013 Exercise"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
```


# Saving Graphs to File

- Don't use the mouse
- Use `ggsave` for ggplot
    - Practice by saving the following plot to file: 

```{r}
ggplot(mtcars, aes(hp, wt)) + 
    geom_point()
#ggsave(FILENAME_HERE, PLOT_OBJECT_HERE)
```

- Base R way: print plots "to screen", sandwiched between `pdf()`/`jpeg()`/`png()`... and `dev.off()`. 
- Vector vs. raster: Images are stored on your computer as either _vector_ or _raster_.
    - __Raster__: an `n` by `m` grid of pixels, each with its own colour. `jpeg`, `png`, `gif`, `bmp`.
    - __Vector__: represented as shapes and lines. `pdf`, [`svg`](https://www.w3schools.com/graphics/svg_intro.asp).
    - For tips: ["10 tips for making your R graphics look their best""](http://blog.revolutionanalytics.com/2009/01/10-tips-for-making-your-r-graphics-look-their-best.html).
    
# Scales; Colour

Scale functions in `ggplot2` take the form `scale_[aesthetic]_[mapping]()`.

Let's first focus on the following plot:

```{r}
p_scales <- ggplot(gapminder, aes(gdpPercap, lifeExp)) +
     geom_point(aes(colour=pop), alpha=0.2)
p_scales + 
    scale_x_log10() +
    scale_colour_continuous(trans="log10")
```

1. Change the y-axis tick mark spacing to 10; change the colour spacing to include all powers of 10.

```{r}
p_scales +
    scale_x_log10() +
    scale_colour_continuous(
        trans  = "log10", 
        breaks = 10^(1:10)
    ) +
    scale_y_continuous(breaks=1:10 * 10) #vector of 1 to 10, multiplied by 10 
#indicates what number we'd like to put tick marks for; if you're out of range, default is to exclude those tick marks
#continuous does not transform the axis at all
```

2. Specify `scales::*_format` in the `labels` argument of a scale function to do the following:
    - Change the x-axis labels to dollar format (use `scales::dollar_format()`)
    - Change the colour labels to comma format (use `scales::comma_format()`)

```{r}
library(scales)
p_scales +
    scale_x_log10(labels=dollar_format()) +
    scale_colour_continuous(
        trans  = "log10", 
        breaks = 10^(1:10),
        labels = comma_format()
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

3. Use `RColorBrewer` to change the colour scheme.
    - Notice the three different types of scales: sequential, diverging, and continuous.
    For discrete or categorical variables, use very different colours. -> continuous
    For a sequence of variables, use a gradient of colours. -> sequential
    For a middle neutral value with a scale of gradients moving away in each direction (i.e. temperature scale to hot and cold), -> diverging

```{r}
## All palettes the come with RColorBrewer:
RColorBrewer::display.brewer.all()
p_scales +
    scale_x_log10(labels=dollar_format()) +
    scale_colour_distiller(  #another one is scale_colour_brewer (one is for continuous, other handles the other two types)
        trans   = "log10",
        breaks  = 10^(1:10),
        labels  = comma_format(),
        palette = "Greens" #get the name from looking at RColorBrewer overview
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

4. Use `viridis` to change the colour to a colour-blind friendly scheme
    - Hint: add `scale_colour_viridis_c` (`c` stands for continuous; `d` discrete).
    - You can choose a palette with `option`.

```{r}
p_scales +
    scale_x_log10(labels=dollar_format()) +
    scale_colour_viridis_c(
        trans   = "log10",
        breaks  = 10^(1:10),
        labels  = comma_format()
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

# Theming

Changing the look of a graphic can be achieved through the `theme()` layer.

There are ["complete themes"](http://ggplot2.tidyverse.org/reference/ggtheme.html) that come with `ggplot2`, my favourite being `theme_bw` (I've grown tired of the default gray background, so `theme_bw` is refreshing).

1. Change the theme of the following plot to `theme_bw()`:

```{r}
ggplot(iris, aes(Sepal.Width, Sepal.Length)) +
     facet_wrap(~ Species) +
     geom_point() +
     labs(x = "Sepal Width",
          y = "Sepal Length",
          title = "Sepal sizes of three plant species") +
    theme_bw() #many possibilities (see tab over)
```

2. Then, change font size of axis labels, and the strip background colour. Others?

```{r}
ggplot(iris, aes(Sepal.Width, Sepal.Length)) +
     facet_wrap(~ Species) +
     geom_point() +
     labs(x = "Sepal Width",
          y = "Sepal Length",
          title = "Sepal sizes of three plant species") +
    theme_bw() +
    theme(axis.text = element_text(size = 16),
          strip.background = element_rect(fill = "orange"),
          panel.background = element_rect(fill = "blue"))

#note: theme elements are not part of the grammer (plot elements are not fundamentally changed)
```


# Plotly

Consider the following plot:

```{r}
(p <- gapminder %>% 
     filter(continent != "Oceania") %>% 
     ggplot(aes(gdpPercap, lifeExp)) +
     geom_point(aes(colour=pop), alpha=0.2) +
     scale_x_log10(labels=dollar_format()) +
     scale_colour_distiller(
         trans   = "log10",
         breaks  = 10^(1:10),
         labels  = comma_format(),
         palette = "Greens"
     ) +
     facet_wrap(~ continent) +
     scale_y_continuous(breaks=10*(1:10)) +
     theme_bw())
```

1. Convert it to a `plotly` object by applying the `ggplotly()` function:

```{r}
#you can make a ggplot object and convert it to plotly (dont need to know plotly syntax)
#install.packages("plotly")

library(plotly)
ggplotly(p)
```

2. You can save a plotly graph locally as an html file. Try saving the above:
    - NOTE: plotly graphs don't seem to show up in Rmd _notebooks_, but they do with Rmd _documents_.

```{r}
p %>% 
    ggplotly() %>% 
    htmlwidgets::saveWidget("LOCATION_GOES_HERE")
```


3. Run this code to see the json format underneath:

```{r}
p %>% 
    ggplotly() %>% 
    plotly_json() #plot is just plain text
```


4. Check out code to make a plotly object from scratch using `plot_ly()` -- scatterplot of gdpPercap vs lifeExp.
    - Check out the [cheat sheet](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf).

```{r}
plot_ly(gapminder, 
        x = ~gdpPercap, #tilda means make a formula with this variable as your formula
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers",
        opacity = 0.2) %>% 
    layout(xaxis = list(type = "log"))
```

5. Add population to form a z-axis for a 3D plot:

```{r}
plot_ly(gapminder, 
        x = ~gdpPercap, 
        y = ~lifeExp, 
        z = ~pop,
        type = "scatter3d",
        mode = "markers",
        opacity = 0.2)
```

#can use gg animate function to animate plots

Why do we make plots? EDA (exploratory data analysis); drawing conclusions from plots -> endpoint
- we've looked at tooling with plots -> intermediate
- grammar and theme are the two aspects that we make decisions about when producing a plot -> intermediate
- effectiveness; what information is contained in the data? -> starting point
- see jenny's animated gif for simplfying data -> does it facilitate comparisons or help you see trends better?
- see Tamara Munsener's book for psychology of colour effects on perception
- don't use pie charts! can be difficult to compare categories (hard to perceive angular distances vs linear, as in a bar plot)
- when you look at your plot, ask yourself what trends can you see; what is hidden; how can you reveal it?
- impossible to make a perfect plot (this is why we have figure captions)


